---
title: "Import Data - ACF - PACF"
author: "Ina Liao"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Setting R code chunk options

The first R code chunk is named **setup**. Here we are setting the options for R code chunks. The choice echo=TRUE means both code and output will appear on report. If for a specific chunk you want different options, you can always change that on the first line as we did in the following chunk. We chose include=FALSE which means that nothing related to this chunk (code and output) will appear on the knitted file.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE) 
```


## Loading packages and initializing

It's useful to designate one code chunk to load packages on the beginning of the file. You can always add to this chunk as needed. But concentrate the packages needed on only one chunk.

```{r package, include=FALSE}
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("forecast")
#install.packages("here")
#install.packages("patchwork")
library(lubridate)
library(ggplot2)
library(forecast) #added for Acf and Pacf functions
library(here)
library(patchwork)
```

## Importing data

For this first example we will import water inflow data for reservoirs in Brazil. We have data for 15 different reservoirs spread all over the country. To import a .txt, .csv or .xlsx file you can use the function *read.table()*. This function will store the data as a data frame and has useful inputs such as

* *file = * : use this input to point to yoru data file. If it's on the same folder as your .Rmd then you only need to write the file name. But if it's on another folder you need to point to the path were file is located;
* *header =* : if your file has a header you should set this to TRUE, o.w. FALSE;
* *skip =* : if your file has rows explaining the data or any other rows on the top that need to be skipped you should just set skip to be equal to the number of row that should be skipped before reading the data. Mote that if header=TRUE, you should not skip the row with the header. The default is *skip=0*;
* *dec =* : define *dec="."* or *dec=","* depending on how it's defined on your set. The default is ".".



```{r Importing files}
#Importing time series data from text file
#I am calling it raw for now because it's data as it is in the file

#Option 1: import using absolute path
#raw_inflow_data <- read.table(file="/Users/lmm89/OneDrive/Duke_University/9_Spring2022/ENV790_TSA/Git_Repo/ENV790_TimeSeriesAnalysis_Sp2022_v2/Data/inflowtimeseries.txt",header=FALSE,skip=0)

#Option 2: import using shortcut "." or ".." or relative path
# the "." will point to the folder set as current directory
# the ".." will point to your parent folder set as current directory

#check which folder is your current directory using function getwd()
getwd()
# my output was "/Users/lmm89/Library/CloudStorage/OneDrive-Personal/Duke_University/11_Spring2023/ENV790_TSA/GitHub/TimeSeriesAnalysis_Sp23" meaning current directory is my project folder, therefore I only need to use "."
# if output was "/Users/lmm89/Library/CloudStorage/OneDrive-Personal/Duke_University/11_Spring2023/ENV790_TSA/GitHub/TimeSeriesAnalysis_Sp23/Lessons" then I would need to use ".."


#read.table not read.csv
raw_inflow_data <- read.table(file="./Data/inflowtimeseries.txt",header=FALSE,skip=0)

#If you want specific columns just trim the table
raw_inflow_data <- raw_inflow_data[,1:5] #the space before the comma means you want all rows
                                 #and 1:5 means all columns from 1 to 5

nhydro <- ncol(raw_inflow_data)-2
nobs <- nrow(raw_inflow_data) 

#If your file does not have header like this one you can add column names after creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3")

#Checking data
head(raw_inflow_data)
str(raw_inflow_data)
```

## Date or time period

The data sets we will work with will be indexed by time, remember we are doing TIME series analysis. After importing your data set make sure that you have your dates right. For this specific inflow file our date is spread in two different columns. The first one is the month and the second the year. The best package to handle date conversion in R is lubridate. Let's see how we can use lubridate functions to combine those two columns into one date object. Note that this is only one example for our particular data set, for more info on lubridate function refer to tho this file [file](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf) and "M1_Intro_DateObjects.Rmd" in our lessons folder.

```{r}
#using package lubridate to paste month and date
#combine column 1 and column 2 and use - to separate year and month 
my_date <- paste(raw_inflow_data[,1],raw_inflow_data[,2],sep="-")

#transform the data type from character into date
my_date <- my(my_date)  #function my from package lubridate
head(my_date) #only contains date data

#add that to inflow_data 
inflow_data <- cbind(my_date,raw_inflow_data[,3:5]) #cbind stands for column bind
head(inflow_data)

#check if there is any missing data
missing_data<-any(is.na(inflow_data))
missing_data
```

## Initial plots

A plot of observed values over time is the first thing you do to start understanding the data set. The default packages on R offer the function *plot()*. But we can generate improved plots using another package *ggplot2*. Package ggplot2 offers better looking plots, additional functionalities, easy incorporation of the date labels, legends, etc.  

```{r Plot}
plot<-ggplot(inflow_data,aes(x=my_date))+
  geom_line(aes(y=HP1),color="gray")+
  geom_line(aes(y=HP2),color="blue")+
  geom_line(aes(y=HP3),color="orange")+
  labs(x="Time",y="Water Inflow (cfs)")
  
plot
```

```{r TEST}
#using package ggplot2
plot_HP1<-ggplot(inflow_data, aes(x=my_date,y=HP1))+
  geom_line()+
  geom_smooth(method=lm,color="blue")+
  labs(x="Time",y="Water inflow (cfs)")
plot_HP1

plot_HP2<-ggplot(inflow_data, aes(x=my_date,y=HP2))+
  geom_line()+
  geom_smooth(method=lm,color="blue")+
  labs(x="Time",y="Water inflow (cfs)")
plot_HP2

plot_HP3<-ggplot(inflow_data, aes(x=my_date,y=HP3))+
  geom_line()+
  geom_smooth(method=lm,color="blue")+
  labs(x="Time",y="Water inflow (cfs)")
plot_HP3

#adding two time series to the same plot
combine_plot<- plot_HP1 / plot_HP2 / plot_HP3
combine_plot
```

In this course we will also use *autoplot()* which is a ggplot() for time series object. But then we will need to convert our inflow data frame to a *ts* object first.


## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. Only the columns with reservoir inflow data should be transformed, not the ones with month and year. Your ts object is not a data frame, but like I said we will need it as a ts for some function we will explore in future scripts.

```{r}
# ts(): to create time-series objects 
# start= c(year, month)
# frequency = observations (when does your data start to repeat)
# yearly data: frequency=1
# monthly data: frequency=12
# daily data: frequency=365

#create a singel vector 
ts_HP1 <- ts(inflow_data[,"HP1"],start=c(1931,1),frequency=12)
ts_HP1
#inflow_data[,"HP1"]: to select all rows in column HP1
ts_HP2 <- ts(inflow_data$HP2,start=c(1931,1),frequency=12)
ts_HP3 <- ts(inflow_data$HP3,start=c(1931,1),frequency=12)
#ts_HP34 <- ts(inflow_data$HP3,start=c(1931,1),frequency=1)

#create a metric
ts_inflow_data<-ts(inflow_data,start=c(1931,1),frequency=12)
ts_inflow_data
```

Note that ts_inflow_data has information on start, end and frequency. We will discuss frequency in future lectures, for now let's just keep the default value for frequency which is 1.

Let's see the plots generated with *autoplot()*.

```{r Autoplot}
autoplot_all<-autoplot(ts_inflow_data) +
  xlab("Time") +
  ylab("Water inflow") +
  labs(color="Reservoir")
autoplot_all
```

```{r}
autoplot_HP1<-autoplot(ts_HP1) +
  xlab("Time") +
  ylab("Water inflow") +
  labs(color="Reservoir")

autoplot_HP2<-autoplot(ts_HP2) +
  xlab("Time") +
  ylab("Water inflow") +
  labs(color="Reservoir")
  
autoplot_HP3<-autoplot(ts_HP3) +
  xlab("Time") +
  ylab("Water inflow") +
  labs(color="Reservoir")

combine_autoplot<-autoplot_HP1/autoplot_HP2/autoplot_HP3
combine_autoplot
```


## Plotting ACF and PACF

There are multiple ways to get acf and pacf plots in R. The default package "stats" have the functions acf() and pacf(). Another option is with the functions Acf() and Pacf() from package "forecast". Package forecast is not a default package. So you need to install and load it on the beginning of your scripts. 

```{r acf}

#ACF and PACF 
# 1. measure the temporal dependency of the stochastic process 
# 2. give us the information about the autoregressive component of the series 

#ACF plot: show the correlation between a time series and lagged versions of itself. ACF can help us understand the seasonal pattern and trend in the dataset. We can build a model that includes lags that show strong correlation between the time series (PACF can help us specify which lag should be added to the model [with the highest y-axis absolute values, closer to -1 or +1])
#anything outside of the blue intervals can be considered as "significant" 

# Option 1: The function acf() and pacf() is from package "stats"
#lag starts at 0
HP1_acf<-stats::acf(ts_HP1)
HP1_acf
#lag in this case will be month cause the frequency is set as 12
#y-axis represents the strength of the correlation 

HP1_pacf<-stats::pacf(ts_HP1)
HP1_pacf
#to remove the effect of the intermediate vairbales 

#Option 2: Computing ACF and PACF with package "forecast"
#lag starts at 1

#The next line will output a plot since we have plot=TRUE and also store the acf values in the object
#if you include lag 2, then we have to include lag 1 and lag 3
HP1_acf_2<-forecast::Acf(ts_HP1,lag.max = 40)
HP1_acf_2

HP1_pacf_2<-forecast::Pacf(ts_HP1)
HP1_pacf_2
```

```{r Test 2}
temp<-c(14,2,16.4,11.9,15.2,18.5,22.1,19.4,25.1,23.4,18.1,22.6,17.2)
ts_temp<-ts(temp)
acf_temp<-Acf(temp)
acf_temp

```

