---
title: "M3: Class Exercise on Trend and Season"
author: "Luana Lima"
output: pdf_document
---

#check the code 


## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)
library(here)
```

## Importing data

Let's continue working with our inflow data for reservoirs in Brazil. 

```{r}

#Importing time series data from text file#
here()
#raw_inflow_data <- read.table(file="../Data/inflowtimeseries.txt",header=FALSE,skip=0)
raw_inflow_data<-read.table(here("Data/inflowtimeseries.txt"),stringsAsFactors = TRUE, header=TRUE,skip=0)


#Trim the table to include only columns you need
nhydro <- ncol(raw_inflow_data)-2
nhydro
nobs <- nrow(raw_inflow_data) 

#If your file does not have header like this one you can add column names after 
#creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3","HP4", "HP5",
                            "HP6","HP7", "HP8","HP9","HP10", "HP11","HP12", 
                            "HP13", "HP14","HP15")

#Checking data
head(raw_inflow_data)
str(raw_inflow_data)
```
## Creating the date object

Here we use the function my() from package lubridate.

```{r}
#using package lubridate
my_date <- paste(raw_inflow_data[,1],raw_inflow_data[,2],sep="-")
my_date <- my(my_date)  #function my from package lubridate
head(my_date)
#add that to inflow_data and store in a new data frame 
inflow_data <- cbind(my_date,raw_inflow_data[,3:(3+nhydro-1)])
head(inflow_data)

```


## Initial Plots

Initial time series plot.

```{r}
#using package ggplot2

#i change from 1 to nhydro
#[1+i]: select the second column (the first column is date)
#ggplot will not work in the for loop without "print"
for(i in 1:nhydro){
  print(ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+i)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+i)],sep="")) 
        )
}

```

## Zeros in the end on data 

The initial plots showed that we have zeros in the end of the data set. It could be missing observation or observation that haven't been observed yet. Use the tail() to find out how many zeros you have and how many lines you will need to remove.

```{r}
#check the final obs on data (for the head and tail, more to come for the missing module)
tail(inflow_data)

#approach to looking for zero within the dataframe 
#using date series dataframe and merge with the exisiting dataframe to find the rows with missing data
```

Note our last observation is from August 2011 but the data file was filled with zeros. Let's remove the last four rows of our data set.

```{r}
#Remove last year by replacing current data frame
inflow_data <- inflow_data[1:(nobs-12),]

#update object with number of observations
nobs <- nobs-12

#Tail again to check if the rows were correctly removed
tail(inflow_data)
```
Fixed!

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_inflow_data <- ts(inflow_data[,2:(2+nhydro-1)],start=c(1931,1),frequency=12) 

#select all column from 2 to 2+15-1 (you can also do nhydro+1)
#to avoid go over from the first to the last columns again

#note that we are only transforming columns with inflow data, not the date columns  #start=my_date[1],end=my_date[nobs],
head(ts_inflow_data,15)

#head() default will show the first 6 rows 
```


## Plotting ACF and PACF

Let's use functions Acf() and Pacf() from package "forecast".

```{r acf}
#Acf and Pacf for HPi
for(i in 1:nhydro){
  par(mfrow=c(1,2))  #place plot side by side
  Acf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep="")) 
  # because I am not storing Acf() into any object, I don't need to specify plot=TRUE 
  Pacf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep=""))
}

#use ACF to see if there is a trend and seasonal component
#we can also use the time series plot to see the trend
#use pacf to identify which lags should be included 
```
```{r Acf and Pacf, message=FALSE}
iHP<-4
plot_grid(
  autoplot(ts_inflow_data[,iHP]),
  autoplot(Acf(ts_inflow_data[,iHP],lags=40,plot=FALSE),main=NULL),
  #if plot=TRUE, another ACF plot will be printed (since autoplot is default to print out the ACF plot)
  nrow=2,ncol=2
)

#after detrend (removing the linear regression), autocorrelation decrease in the detrended series 
#purpose: detrend to let time representative for the needed time (include the needed variables in the model)
#using ACF to understand which time frame (lags #) should be included in the model (which lag is less dependent on the time) 
#capture the model util lag 20 

#use ACF or PACF plot?
#use ACF to see the trend and the seasonal components 
#what is the purpose of pacf plot??
```

## Trend Component - Linear Model

Let's identify and remove trend component. You start by fitting a linear model to $Y_t=\beta_0+\beta_1*t+\epsilon_t$. 

```{r}
#Create vector t to store the number of rows 
#we could also use my_date but it's more complicated 
t<-1:nobs 

iHP <-4
#to select the HP4
#the first column is the date column 
#iHP+1: the fifth column

#Fit a linear trend to TS of iHP 
#use lm to find the beta0 and beta1
linear_trend<-lm(inflow_data[,iHP+1]~t)
summary(linear_trend)

#there is a strong relationship between inflow in HP4 and time
```


```{r}
#remove the trend from series
#use the $ to find the objects stored in the object 
beta0<-linear_trend$coefficients[1]
beta1<-linear_trend$coefficients[2]

y_detrend<-inflow_data[,iHP+1] - (beta0 + beta1*t)
#subtract the series that we believe has trend components
y_detrend

#create another data frame
df_detrend<-data.frame("date"=inflow_data$my_date,
                       "observed"=inflow_data[,iHP+1],
                       "detrend"=y_detrend)

#autoplot and autolayer will only work with the time series object 
#still has the variability (might have other variables that is affecting the time series data)
#how to confirm if the detrend the series is correct 
#if the trend is non-linear moving -> using moving average 
```

Note that blue line is our original series, red line is our trend, green line is our original series minus the trend or in other words the detrended series. And in orange is the trend line for the detrended series which has slope 0 meaning we were able to effectively eliminate the trend with a linear model.


## Seasonal Component

Now let's shift attention to the seasonal component.

```{r}
#Use seasonal means model by assigning dummy variable
iHP<-1

#create dummies
seasonal_dummies<-seasonaldummy(ts_inflow_data[,iHP])

#seasonal dummy only work when the seasonal frequency is specified (frequency=12)
seasonal_linear<-lm(inflow_data[,(iHP+1)]~seasonal_dummies) #use the original data instead of the time series (ideally we have to use the detrended series)
summary(seasonal_linear)

seasonal_linear$coefficients
# 12 coefficients for each month 

#store coeff
slope<-seasonal_linear$coefficients[2:12]
intercept<-seasonal_linear$coefficients[1]

#create seasonal components 
# %*%: calculate element by element (slope multiply Jan/Feb...)
seasonal_components<-array(0,nobs)

for i in (1:nobs){
  seasonal_components[i]<-intercept+slope %*% seasonal_dummies[i,]
  #the month is stored in rows in seasonal dummies 
}


```

##Exercise

Fit trend and seasonal for the other variables HP2, HP3, ...


## Stationarity Tests in R

Some test only work for non-seasonal data. So let's create another series with yearly averages for inflow.

```{r}
#Group data in yearly steps instances


```

### Mann Kendall 

Check for deterministic trend.

```{r}

#Since I have seasonal data I cannot use the simple MannKendall()
#another example of functions that need a ts object


#Use yearly date to run Mann Kendall

```

### Spearman Correlation test 

Also check for deterministic trend, for non-seasonal data only.

```{r}
#Deterministic trend with Spearman Correlation Test
print("Results from Spearman Correlation")


#with cor.test you can get test statistics

```

### Augmented Dickey Fuller (ADF)

Used to detect stochastic trend.

```{r}
#Null hypothesis is that data has a unit root
print("Results for ADF test/n")

#Now let's try the yearly data
print("Results for ADF test on yearly data/n")

```

##Exercise

Run the stationarity tests for the other variables HP2, HP3, ...

