---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2024"
subtitle: "Assignment 6 - Due date 02/28/24"
author: "Ina Liao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  markdown: 
    wrap: 72
---

## Directions

You should open the .rmd file corresponding to this assignment on
RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you
will do is rename the file such that it includes your first and last
name (e.g., "LuanaLima_TSA_A06_Sp24.Rmd"). Then change "Student Name" on
line 4 with your name.

Then you will start working through the assignment by **creating code
and output** that answer each question. Be sure to use this assignment
document. Your report should contain the answer to each question and any
plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into
a single PDF file. Submit this pdf using Sakai.

R packages needed for this assignment: "ggplot2", "forecast", "tseries"
and "sarima". Install these packages, if you haven't done yet. Do not
forget to load them before running your script, since they are NOT
default packages.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r Install Packages, warning=FALSE,message=FALSE}
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(tidyverse)  
library(here)
library(knitr)
library(ggthemes)
library(cowplot)
library(dplyr)
#install.packages("sarima")
library(sarima)
```

This assignment has general questions about ARIMA Models.

## Q1

Describe the important characteristics of the sample autocorrelation
function (ACF) plot and the partial sample autocorrelation function
(PACF) plot for the following models:

-   AR(2)

> Answer: The ACF plot shows an exponential decay in autocorrelation.
> PACF plot has the clear cutoff at lag 1, which can help us identify
> the order of the MA model.

-   MA(1)

> Answer: The ACf plot can help us identify the order of the MA model,
> which will have the clear cutoff at lag 1. The PACF plot displays an
> exponential decay in partial autocorrelation.

## Q2

Recall that the non-seasonal ARIMA is described by three parameters
ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component,
$d$ is the number of times the series need to be differenced to obtain
stationarity and $q$ is the order of the moving average component. If we
don't need to difference the series, we don't need to specify the "I"
part and we can use the short version, i.e., the ARMA$(p,q)$.

(a) Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with
    parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR
    coefficient and the $\theta$ refers to the MA coefficient. Use the
    `arima.sim()` function in R to generate $n=100$ observations from
    each of these three models. Then, using `autoplot()` plot the
    generated series in three separate graphs.

```{r ARMA Series 1}
#set the parameters
phi<-0.6 #AR coefficient
theta<-0.9 #MA coefficient
n<-100

#generate data from ARMA(p,q)
#in the arima.sim function, the argument order=c(p,d,q) is not needed
#beacuse through assigning the cofficient, the arima.sim will know if it is AR or MA process
arma_10<-arima.sim(model=list(ar=phi),n=n)
arma_01<-arima.sim(model=list(ma=theta),n=n)
arma_11<-arima.sim(model=list(ar=phi,ma=theta),n=n)

#plot the series
plot_10<-autoplot(arma_10)
plot_10
plot_01<-autoplot(arma_01)
plot_01
plot_11<-autoplot(arma_11)
plot_11
```

(b) Plot the sample ACF for each of these models in one window to
    facilitate comparison (Hint: use `cowplot::plot_grid()`).

```{r ARMA_ACF}
plot_combine_acf<-cowplot::plot_grid(
  autoplot(Acf(arma_10,lag.max=40,plot=FALSE)),
  autoplot(Acf(arma_01,lag.max=40,plot=FALSE)),
  autoplot(Acf(arma_11,lag.max=40,plot=FALSE)),
  ncol=3
)
plot_combine_acf
```

(c) Plot the sample PACF for each of these models in one window to
    facilitate comparison.

```{r ARMA_PACF}
plot_combine_pacf<-cowplot::plot_grid(
  autoplot(Pacf(arma_10,lag.max=40,plot=FALSE)),
  autoplot(Pacf(arma_01,lag.max=40,plot=FALSE)),
  autoplot(Pacf(arma_11,lag.max=40,plot=FALSE)),
  ncol=3
)
plot_combine_pacf
```

(d) Look at the ACFs and PACFs. Imagine you had these plots for a data
    set and you were asked to identify the model, i.e., is it AR, MA or
    ARMA and the order of each component. Would you be able identify
    them correctly? Explain your answer.

> Answer: We can see an exponential decay in the ACF plot, and a sharp
> cutoff at lag-1 in the PACF plot, which suggests the AR model. The ACF
> plot shows a sharp cutoff at lag-1 and the PACF plot show an
> exponential decay, which suggests the MA model. We can observe an
> exponential decay in ACF and PACF plots, which suggests the ARMA
> model. Relying solely on PACF and ACF plots to determine the model may
> be insufficient, as the patterns in these plots can often overlap and
> be difficult to interpret without additional information. \# the
> pattern in the ACF plot and PACf plot seem to be overlapped in the
> ARMA(1,1)

(e) Compare the PACF values R computed with the values you provided for
    the lag 1 correlation coefficient, i.e., does $\phi=0.6$ match what
    you see on PACF for ARMA(1,0), and ARMA(1,1)? Should they match?

> Answer: In ARMA(1,0) and ARMA(1,1), the PACF values at lag-1 match the
> AR coefficient (phi=0.6). If there is no randomness in the stimulated
> data, the partial autocoefficient should match the specified
> coefficient.

(f) Increase number of observations to $n=1000$ and repeat parts
    (b)-(e).

```{r ARMA Series 2}
#set the parameters
phi<-0.6 #AR coefficient
theta<-0.9 #MA coefficient
n<-1000

#generate data from ARMA(p,q)
arma_10<-arima.sim(model=list(ar=phi),n=n)
arma_01<-arima.sim(model=list(ma=theta),n=n)
arma_11<-arima.sim(model=list(ar=phi,ma=theta),n=n)

#plot the series
plot_10<-autoplot(arma_10)
plot_10
plot_01<-autoplot(arma_01)
plot_01
plot_11<-autoplot(arma_11)
plot_11

#ACF plot
plot_combine_acf<-cowplot::plot_grid(
  autoplot(Acf(arma_10,lag.max=40,plot=FALSE)),
  autoplot(Acf(arma_01,lag.max=40,plot=FALSE)),
  autoplot(Acf(arma_11,lag.max=40,plot=FALSE)),
  ncol=3
)
plot_combine_acf

#PACF plot
plot_combine_pacf<-cowplot::plot_grid(
  autoplot(Pacf(arma_10,lag.max=40,plot=FALSE)),
  autoplot(Pacf(arma_01,lag.max=40,plot=FALSE)),
  autoplot(Pacf(arma_11,lag.max=40,plot=FALSE)),
  ncol=3
)
plot_combine_pacf
```

> Answer: We can see an exponential decay in the ACF plot, and a sharp
> cutoff at lag-1 in the PACF plot, which suggests the AR model. The ACF
> plot shows a sharp cutoff at lag-1 and the PACF plot show an
> exponential decay, which suggests the MA model. We can observe an
> exponential decay in ACF and PACF plots, which suggests the ARMA
> model. Relying solely on PACF and ACF plots to determine the model may
> be insufficient, as the patterns in these plots can often overlap and
> be difficult to interpret without additional analysis. #beacuse
> ARMA(0,1) and ARMA(1,1) include the moving average components, the
> autocoefficient in the PACF plot do not match the AR coefficient
> (phi=0.6) #we can also ask R to print out the PACF value instead of
> just looking at the value in the plot

> Answer: In ARMA(1,0) and ARMA(1,1), the PACF values at lag-1 do not
> match the AR coefficient (phi=0.6), which might due to the randomness
> in the stimulated data.

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$,
    i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the
    equation.

<!-- -->

1.  p is the number of autoregressive term. p=1 because of $y_{t-1}$.
2.  d is the number of differences. d=0 because there is no differencing
    term applied to $y_{t}$.
3.  q is the number of moving average term. q=1 because of
    $0.1*a_{t-1}$.
4.  P is the seasonal order of autoregressive term. P=1 because of
    $y_{t-1}$.
5.  D is the seasonal order of differences.
6.  Q=0 because there is no MA terms.
7.  s is the length of the seasonal cycle. s=12 because $y_{t-12}$

ARIMA(1,d,1)x(1,D,0)12

$y_{t}$. depends on two previous observation $y_{t-1}$, and $y_{t-12}$,
since 12 is the seasonal lag the order of the AR,the coefficient of the
AR component will be p=1 and the SAM component will be P=1. The value of
$y_{t}$ also depends on the previous residuals $a_{t-1}$ meaning we have
a MA components of order q=1. There is no $a_{t-1}$ so SAM components
order will be Q=1.

With respect to differencing, it is hard to tell from the equation since
there is no constant term, therefore we are either working with a zero
mean process or the series has been differenced. Given the constraint of
the SARIMA model that d+D\<= 2, the d and D values might be the
following combinations: 1. d=1, and D=0 2. d=0, and D=1 3. d=0, and D=0
4. d=1, and D=1

(b) Also from the equation what are the values of the parameters, i.e.,
    model coefficients.

<!-- -->

1.  The autoregressive coefficient $\phi=0.7$.
2.  The moving average coefficient $\theta=-0.1$.
3.  The seasonal autoregressive coefficient $\Phi=-0.25$. The seasonal
    moving average coefficient $\Theta=0$.

## Q4

Simulate a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with
$\phi =0 .8$ and $\theta = 0.5$ using the `sim_sarima()` function from
package `sarima`. The $12$ after the bracket tells you that $s=12$,
i.e., the seasonal lag is 12, suggesting monthly data whose behavior is
repeated every 12 months. You can generate as many observations as you
like. Note the Integrated part was omitted. It means the series do not
need differencing, therefore $d=D=0$. Plot the generated series using
`autoplot()`. Does it look seasonal?

```{r ARIMA}
#set the parameters
phi<-0.8 #AR coefficient
theta<-0.5 #MA coefficient
#integrated part was omitted so the series do not need differencing

#define ARIMA model
sarima<-sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000) 

#convert to time series object
#arima.sim will generate time series object 
#sarima will generate numeric results
#create time series object 
ts_sarima<-ts(sarima,frequency=12)

#plot the series
plot_sarima<-autoplot(ts_sarima)
plot_sarima

```

The plot show a wave-like pattern, which suggests possible presence of a
seasonal component in the series.

## Q5

Plot ACF and PACF of the simulated series in Q4. Comment if the plots
are well representing the model you simulated, i.e., would you be able
to identify the order of both non-seasonal and seasonal components from
the plots? Explain.

```{r}
plot_sarima_acf_pacf<-cowplot::plot_grid(
  autoplot(Acf(ts_sarima,lag.max=40,plot=FALSE)),
  autoplot(Pacf(ts_sarima,lag.max=40,plot=FALSE)),
  ncol=2
)
plot_sarima_acf_pacf
```

There are spikes at lags 12, 24, and 36 on the ACF plot, and a spike at
lag 12 on the PACF plot. These spikes suggest seasonal components might
exist in the time series data. Further statistical testing is required
to confirm the existence of a seasonal component.
