---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Ina Liao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r Setup, include=FALSE} 
knitr:::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=FALSE,fig.align = "center", dev = "cairo_pdf", fig.pos = "H")
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r Install Packages,message=FALSE}
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("forecast")
#install.packages("here")
#install.packages("patchwork")
#install.packages("tidyr")
#install.packages("knitr")
#install.packages("kableExtra")
#install.packages("ggthemes")
#install.packages("cowplot")
library(lubridate)
library(ggplot2)
library(forecast) #added for Acf and Pacf functions
library(here)
library(patchwork)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggthemes)
library(cowplot)
```

```{r Import data, echo=TRUE, results = "hide"}
#check working directory
here()

#import data
raw_energy<-read.csv(here("Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv"),stringsAsFactors = TRUE,skip=0,header=TRUE)
raw_energy

#transform date format
Date<-ym(raw_energy$Month)
raw_energy<-cbind(Date,raw_energy[,2:14])
raw_energy
```

```{r Select Needed Data}
df_energy<-raw_energy[,c(1,5,6)]
head(df_energy)

#rename column names 
new_names<-c("Date","Renewable Production","Hydroelectric Consumption")
colnames(df_energy)<-new_names
head(df_energy)

#check if there is any missing data
missing_data<-any(is.na(df_energy))
missing_data #there is no missing data in the dataframe
```

```{r Time Series Objects}
#find the start date
year1<-year(df_energy$Date[1])
month1<-month(df_energy$Date[1])

ts_energy<-ts(df_energy[,2:3],start=c(year1,month1),frequency=12) 
```

##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r Set Plot Theme}
my_plot_theme<- theme_calc()+
  theme(
    #plot background
    plot.background = element_rect(color="gray"),
    
    #plot title
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    
    #gridlines 
    panel.grid.major=element_line("white"),
    panel.grid.minor = element_line("white"),
    axis.ticks=element_line(color="gray"),
    
    #legend 
    legend.key=element_rect(color="white"),
    legend.background = element_rect(color="white"),
    legend.text = element_text(size = 8),
    legend.position="right"
      )
theme_set(my_plot_theme)
```

```{r Number of rows and columns}
#for the purpose of the for loop
num_row<-nrow(df_energy)
num_col<-ncol(df_energy) #the first column is Date
```

```{r Plots}
#place plot side by side
par(mfrow=c(1,3)) 

for(i in 1:2){ 
plot(df_energy$Date,ts_energy[,i],type="l",col="black",
ylab="Energy (Trillion Btu)",xlab=NA,main=paste(colnames(df_energy[i]),sep=""))
Acf(ts_energy[,i],lag.max=40,main=paste("ACF for",colnames(df_energy[i]),sep=" "))
Pacf(ts_energy[,i],lag.max=40,main=paste("PACF for",colnames(df_energy[i]),sep=" "))
}

#use "plot" in the for loop instead of print(ggplot)
```


### Q2
From the plot in Q1, do the series Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Renewable energy production has been increasing over time, but there is no clear seasonal variation in the trend. Hydroelectric power consumption does not show any noticeable growth or decline over time, but it does have seasonal components in the data. 

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r Regression_Renewable}
#create a time vector 
t<-c(1:num_row)

#create a new data frame
#use the time series object
df_renew<-data.frame("time"=t,"renewable"=ts_energy[,1])

#run linear regression 
linear_renew=lm(ts_energy[,1]~t,df_renew)
#use the time series object
summary(linear_renew)

#store the intercept and slope
intercept_renew=as.numeric(linear_renew$coefficients[1])
slope_renew=as.numeric(linear_renew$coefficients[2])
```

```{r Regression_Hydro}
#create a new data frame
df_hydro<-data.frame("time"=t,"hydro"==ts_energy[,2])

#run linear regression 
linear_hydro=lm(ts_energy[,2]~t,df_hydro)
summary(linear_hydro)

#store the intercept and slope
intercept_hydro=as.numeric(linear_hydro$coefficients[1])
slope_hydro=as.numeric(linear_hydro$coefficients[2])
```

Regression results show a slight decreasing linear trend with a slope of -0.00985.

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r Detrend_Renewable}
#create the detrended time series from linear trend 
detrend_renew<-ts_energy[,1]-(slope_renew*df_renew$time+intercept_renew)
#use the time series object

#create detrended time series object 
ts_detrend_renew<-ts(detrend_renew,start=c(year1,month1),frequency=12)
```

```{r Detrend_Hydro}
#create the detrended time series from linear trend 
detrend_hydro<-ts_energy[,2]-
  (slope_hydro*df_hydro$time+intercept_hydro)
#use the time series object

#create detrended time series object 
ts_detrend_hydro<-ts(detrend_hydro,start=c(year1,month1),frequency=12)
```

```{r Detrend_all,echo=TRUE, results = "hide",message=FALSE}
#combine detrended time series object 
ts_detrend_all<-cbind(ts_detrend_renew,ts_detrend_hydro)
```

```{r Plot_Trend_Detrend_1}
plot_detrend_renew<-autoplot(ts_energy[,1],series="Time Series")+
  autolayer(ts_detrend_all[,1], series="Detrended Series")+
  labs(x="Time",y="Renewable Production (quadrillion Btu)",color="")
plot_detrend_renew

plot_detrend_hydro<-autoplot(ts_energy[,2],series="Time Series")+
  autolayer(ts_detrend_all[,2], series="Detrended Series")+
  labs(x="Time",y="Hydroelectrical Power Consumption (quadrillion Btu)",color="")
plot_detrend_hydro
```

After removing linear trends, detrended renewable energy production show less upward trend over time, which might because of the underlying trend in renewable energy production being approximately linear. Both original and detrended hydroelectric power consumption do not show any significant increase or decrease over the time period. 


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?

```{r Plot_Trend_Detrend_2}
#place the plot side by side 
par(mfrow=c(2,3))

for(i in 1:2){ 
  #original time series
  plot(df_energy$Date,ts_energy[,i],type="l",col="black",
       ylab="Energy (Trillion Btu)",xlab=NA,
       main=paste(colnames(df_energy[i+1]),sep=""))
  Acf(ts_energy[,i],lag.max=40,
      main=paste("ACF for",colnames(df_energy[i+1]),sep=" "))
  Pacf(ts_energy[,i],lag.max=40,
       main=paste("PACF for",colnames(df_energy[i+1]),sep=" "))
  
  #detrended time series
  plot(df_energy$Date,ts_detrend_all[,i],type="l",col="black",
       ylab="Energy (Trillion Btu)",xlab=NA,
       main=paste(colnames(df_energy[i+1]),sep=""))
  Acf(ts_detrend_all[,i],lag.max=40,
      main=paste("ACF for",colnames(df_energy[i+1]),sep=" "))
  Pacf(ts_detrend_all[,i],lag.max=40,
       main=paste("PACF for",colnames(df_energy[i+1]),sep=" "))
}
```

```{r Plot_Detrend, ACF and PACF}
plot_grid(
  autoplot(Acf(ts_energy[,1],lag.max=40,plot=FALSE),main="Renewable Original"),
  autoplot(Acf(ts_detrend_all[,1],lag.max=40,plot=FALSE),main="Renewable Detrended"),
  autoplot(Pacf(ts_energy[,2],lag.max=40,plot=FALSE),main=" ",),
  autoplot(Pacf(ts_detrend_all[,2],lag.max=40,plot=FALSE),main=" "),
  nrow=2,ncol=2
)

```
The autocorrelation was not significantly affected by removing trends in renewable energy production and hydroelectric power consumption. This may becasue of (1) the presence of nonlinear trends in the data, which cannot be removed by a linear detrending process; and (2) the presence of seasonal components in the data. 


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

From the time series and ACF plot, we can not see a significant seasonal trend in the renewable energy production. However, hydroelectric power consumption shows a significant seasonal trend.

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r Seasonal Mean_Renewable}
#create the seasonal dummies 
seasonal_dummies_renew<-seasonaldummy(ts_energy[,1])
#use the original time series object here since the purpose is to identify the seasonal trend in the original series

#fit the detrended series to the seasonal dummies
seasonal_mean_renew=lm(ts_energy[,1]~seasonal_dummies_renew)
summary(seasonal_mean_renew)

#store regression coefficients
slope_renew_deseason=seasonal_mean_renew$coefficients[2:12]
intercept_renew_deseason=seasonal_mean_renew$coefficients[1]
```

```{r Seasonal Mean_Hydro}
#create the seasonal dummies 
seasonal_dummies_hydro<-seasonaldummy(ts_energy[,2])

#fit the detrended series to the seasonal dummies
seasonal_mean_hydro=lm(ts_energy[,2]~seasonal_dummies_hydro)
summary(seasonal_mean_hydro)

#store regression coefficients
slope_hydro_deseason=seasonal_mean_hydro$coefficients[2:12]
intercept_hydro_deseason=seasonal_mean_hydro$coefficients[1]
```

Hydroelectric power consumption show a clear seasonal trend. The null hypothesis assumes the absence of seasonality, while the alternative hypothesis proposes its presence. Based on the results of F-test, we have have sufficient evidence to reject the null hypothesis for hydroelectric power consumption (p-value<0.05), indicating the existence of seasonality. However, we do not have sufficient evidence to reject the null hypothesis for renewable energy production (p-value>0.05).

Since p-value is compared with 0.05 instead of 0.001, do not use p-value<0.001 to express p-value<0.05
 
### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r Deseason_Renewable}
#compute seasonal components 
seasonal_components_renew=array(0,num_row)
for (i in 1:num_row){
  seasonal_components_renew[i]=ts_energy[i,1]-
    (intercept_renew_deseason+slope_renew_deseason%*%seasonal_dummies_renew[i,])
}
#the symbol %*% means inner product

#transform seasonal components into a time series object
ts_deseason_renew<-ts(seasonal_components_renew,start=c(year1,month1),frequency=12)
```

```{r Deseason_Hydro}
#compute seasonal components 
seasonal_components_hydro=array(0,num_row)
for (i in 1:num_row){
  seasonal_components_hydro[i]=ts_energy[i,2]-
    (intercept_hydro_deseason+slope_hydro_deseason%*%seasonal_dummies_renew[i,])
}

#transform seasonal components into a time series object
ts_deseason_hydro<-ts(seasonal_components_hydro,start=c(year1,month1),frequency=12)
```

```{r Deseason_all}
ts_deseason_all<-cbind(ts_deseason_renew,ts_deseason_hydro)
```

```{r Plot_Trend_Deseason_1}
plot_deseason_renew<-autoplot(ts_energy[,1],series="Time Series")+
  autolayer(ts_deseason_all[,1], series="Deseason Series")+
  labs(x="Time",y="Renewable production (quadrillion Btu)",color="")
plot_deseason_renew

plot_deseason_hydro<-autoplot(ts_energy[,2],series="Time Series")+
  autolayer(ts_deseason_all[,2], series="Deseason Series")+
  labs(x="Time",y="Hydroelectric Consumption (quadrillion Btu)",color="")
plot_deseason_hydro

```

After removing seasonal components, both renewable energy production and hydroelectric power consumption show wave-like patterns with equally spaced peaks and troughs.
 

### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?

```{r Plot_Trend_Deseason_ACF_PACF}
#place the plot side by side 
par(mfrow=c(2,3))

for(i in 1:2){ 
  #original time series
  plot(df_energy$Date,ts_energy[,i],type="l",col="black",
       ylab="Energy (Trillion Btu)",xlab=NA,
       main=paste(colnames(df_energy[i+1]),sep=""))
  Acf(ts_energy[,i],lag.max=40,
      main=paste("ACF for",colnames(df_energy[i+1]),sep=" "))
  Pacf(ts_energy[,i],lag.max=40,
       main=paste("PACF for",colnames(df_energy[i+1]),sep=" "))
  
  #detrended time series
  plot(df_energy$Date,ts_deseason_all[,i],type="l",col="black",
       ylab="Energy (Trillion Btu)",xlab=NA,
       main=paste(colnames(df_energy[i+1]),sep=""))
  Acf(ts_deseason_all[,i],lag.max=40,
      main=paste0("ACF for",colnames(df_energy[i+1]),sep=" "))
  Pacf(ts_deseason_all[,i],lag.max=40,
       main=paste("PACF for",colnames(df_energy[i+1]),sep=" "))
}

```

After removing the seasonal components from the renewable energy production time series data, we observed spikes at lags 13, 24, and 36 in the ACF plot. These spikes suggest that there is no clear seasonality in the data. In the hydroelectric power consumption data, we can see a clearer seasonality from the ACF plot after removing seasonal components in the data. 
Both renewable energy production and hydroelectric consumption have a clear cutoff at spike 13 in the PACF plot, we can thus further build a time series model that includes lags 1 and 13 to better understand the variability of the data. 
