---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 4 - Due date 02/12/24"
author: "Ina Liao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r Setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r Install Packages, message=FALSE,warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)
library(here)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggthemes)
library(cowplot)
library(dplyr)
library(base)
```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 

```{r Import data, echo=TRUE, results = "hide"}
#check working directory
here()

#import data
raw_energy<-read.csv(here("Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv"),stringsAsFactors = TRUE,skip=0,header=TRUE)
raw_energy

#transform date format
Date<-ym(raw_energy$Month)
raw_energy<-cbind(Date,raw_energy[,2:14])
raw_energy
```

```{r Select Needed Data}
df_energy<-raw_energy[,c(1,5)]
head(df_energy)

#rename column names 
new_names<-c("Date","Renewable Production")
colnames(df_energy)<-new_names
head(df_energy)
```

```{r Time Series Objects}
#find the start date
year1<-year(df_energy$Date[1])
month1<-month(df_energy$Date[1])

#time series object
ts_energy<-ts(df_energy,start=c(year1,month1),frequency=12) 
```

```{r Set Plot Theme}
my_plot_theme<- theme(
    #plot title
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10,angle = 90, vjust = 0.5, hjust = 0.5),

    #legend 
    legend.text = element_text(size = 10),
    legend.position="right"
      )
theme_set(my_plot_theme)
```


## Stochastic Trend and Stationarity Tests

### Q1
Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series Do the series still seem to have trend?

```{r Differencing}
ts_energy_diff<-diff(ts_energy[,2],lag = 1,differences = 1)
#diff can used to eliminated seasonality by differences at the seasonal level

plot_time_diff <- autoplot(ts_energy[,2],series="Time Series")+
  autolayer(ts_energy_diff, series="Differenced Series")+
  labs(x="Time",y="Production (quadrillion Btu)",title="Renewable Energy Production",color="")
plot_time_diff

#use scale_color_manual() to change the color in autoplot 
 
#the renewable production is stochastic trend 
#use differencing method to detrend instead of linear detrending method
```
From the time series plot, it seems that the trend level has been decreased in the differenced time series.

### Q2
Copy and paste part of your code for A3 where you run the regression for Total Renewable Energy Production and subtract that from the orinal series. This should be the code for Q3 and Q4. make sure you use the same name for you time series object that you had in A3.

```{r Detrend}
#create a time vector 
num_row<-nrow(df_energy)
t<-c(1:num_row)

#create a new data frame
df_renew<-data.frame("time"=t,"renewable"=df_energy$`Renewable Production`)

#run linear regression 
linear_renew=lm(renewable~t,df_renew)
summary(linear_renew)

#store the intercept and slope
intercept_renew=as.numeric(linear_renew$coefficients[1])
slope_renew=as.numeric(linear_renew$coefficients[2])

#create the detrended time series from linear trend 
detrend_renew<-df_energy$`Renewable Production`-
  (slope_renew*df_renew$time+intercept_renew)

#create detrended time series object 
ts_detrend_renew<-ts(detrend_renew,start=c(year1,month1),frequency=12)
```



### Q3
Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in Q2 using linear regression. 

Using autoplot() + autolayer() create a plot that shows the three series together. Make sure your plot has a legend. The easiest way to do it is by adding the `series=` argument to each autoplot and autolayer function. Look at the key for A03 for an example. 

```{r Plot: Time Series}
plot_all<-autoplot(ts_energy[,2],series="Time Series")+
  autolayer(ts_energy_diff,series="Differenced Series")+
  autolayer(ts_detrend_renew,series="Detrended Series")+
  labs(x="Time",y="Production (quadrillion Btu)",title="Renewable Energy Production",color="")
plot_all
```

### Q4
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the autoplot() or Acf() function - whichever you are using to generate the plots - to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r Plot: ACF}
#place the plot side by side 
par(mfrow=c(1,3))
df_all<-cbind(ts_energy[,2],ts_detrend_renew,ts_energy_diff)

#rename column names 
new_names<-c("Original time series","Detrended time series","Differenced time series")
colnames(df_all)<-new_names
head(df_all)

for(i in 1:3){ 
  Acf(df_all[,i],lag.max=40,main=paste("ACF for",new_names[i]),
      ylim=c(-0.5,1))
}

#df_all[i] will not print the column names
#correlation is still high in the higher lag 
#the seasonal component is clearer in the ACF plot with the differenced time series (spikes at 12, 24, and 36)
```

The differencing method appears to be a more effective way of removing the trend. 
(1) After using linear regression to detrend, the autocorrelation remains high, which indicates that the trend may not be linear and still exists in the detrended time series.
(2) After taking the difference, we can see that there are several spikes in the differenced time series. With this information, we can further identify months with high autocorrelation.


### Q5
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. What is the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r Seasonal Mann-Kendall and ADF Test}
#Seasonal Mann-Kendall
SMK_original_ts<-SeasonalMannKendall(ts_energy[,2])
print(summary(SMK_original_ts))

#use the smk.test in the trend package 
library(trend)
summary(trend::smk.test(ts_energy[,2]))
#use either summary or print to show the result 
#for ADF, we have to use print to see the p-value 
#the significant is the same across different seasons

#ADF test
ADF_original_ts<-adf.test(ts_energy[,2],alternative = "stationary")
print(ADF_original_ts)

#MannKendall is used to determine if the series has determined trend 
#ADF test is used to determine if the series has stochastic trend 
#both tests are needed since they are testing for different trends  
```

(1) We have sufficient evidence to reject the null hypothesis that renewable energy is stationary (p-value < 0.05). The renewable energy production has a strong time dependence (tau = 0.783) and might have an increasing trend over time (S score > 0).
(2) We do not have sufficient evidence to reject the null hypothesis that there is a unit root in renewable energy production time series  (p-value > 0.05). The result from the Augmented Dickey-Fuller Test suggests that there might be a stochastic trend in the renewable energy production.
(3) To sum up, there is a non-linear trend in renewable energy production, consistent with previous observations.


### Q6
Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. Convert the accumulates yearly series into a time series object and plot the series using autoplot().

```{r Yearly Series}
#store series in a matrix
#row: month, column: year
energy_matrix<-matrix(ts_energy[,2],byrow=FALSE,nrow=12)

#calculate production annual mean
energy_yearly<-colMeans(energy_matrix)

#NA will be removed when calculating the average values 

#create another column for year
date_new<-ymd(df_energy$Date)
year<-c(year(first(date_new)):year(last(date_new)))

#merge as a dataframe
df_energy_yearly<-data.frame(year, energy_yearly)
head(df_energy_yearly)

#create time series object
ts_energy_yearly<-ts(df_energy_yearly[,2],start=1973,end=2023,frequency=1) 
head(ts_energy_yearly)

#plot
plot_yearly<-autoplot(ts_energy_yearly)+
labs(x="Year",y="Production (quadrillion Btu)",title="Renewable Energy Production",color="")
plot_yearly
```

### Q7
Apply the Mann Kendal, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the monthly series, i.e., results for Q6?

```{r Stationary Test}
#Mann-Kendall
MK_yearly_ts<-MannKendall(ts_energy_yearly)
print(summary(MK_yearly_ts))

#Spearman correlation rank test
Spearman_yearly_ts<-cor.test(ts_energy_yearly,year,method="spearman")
print(Spearman_yearly_ts)

#ADF test
ADF_yearly_ts<-adf.test(ts_energy_yearly,alternative = "stationary")
print(ADF_yearly_ts)
```

(1) We have sufficient evidence to reject the null hypothesis that renewable energy is stationary (p-value < 0.05). The renewable energy production has a strong time dependence (tau = 0.799) and might have an increasing trend over time (S score > 0).
(2) We have sufficient evidence to reject the null hypothesis that renewable energy is stationary (p-value < 0.05).The renewable energy production has a strong time dependence (rho = 0.913). 
(3) We do not have sufficient evidence to reject the null hypothesis that there is a unit root in renewable energy production time series  (p-value > 0.05). The result from the Augmented Dickey-Fuller Test suggests that there might be a stochastic trend in the renewable energy production.
The test results align with the test results from the previous questions.

